---
title: 'CMGFM: simulation'
author: "Wei Liu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{CMGFM: simulation}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the usage of CMGFM for the analysis of high-dimensional multimodality data with additional  covariates, by comparison with other methods. 


The package can be loaded with the command:
```{r  eval = TRUE}
library(CMGFM)
```

## Generate the simulated data
First, we generate the data simulated data. 
```{r  eval = TRUE}
pveclist <- list('gaussian'=c(50, 150),'poisson'=c(50, 150),
                 'binomial'=c(100,60))
q <- 6
sigmavec <- rep(1,3)
pvec <- unlist(pveclist)
methodNames <- c("CMGFM", "GFM", "MRRR", "LFR" , "GPCA", "COAP")

datlist <- gendata_cmgfm(pveclist = pveclist, seed = 1, n = 300,d = 3,
                           q = q, rho = rep(1,length(pveclist)), rho_z=0.2,
                           sigmavec=sigmavec, sigma_eps=1)
str(datlist)
XList <- datlist$XList
Z <- datlist$Z
numvarmat <- datlist$numvarmat
Uplist <- list()
k <- 1
for(id in 1:length(datlist$Uplist)){
    for(im in 1:length(datlist$Uplist[[id]])){
      Uplist[[k]] <- datlist$Uplist[[id]][[im]]
      k <- k + 1
    }
}
types <- datlist$types
```
Fit the CMGFM model using the function `CMGFM()` in the R package `CMGFM`. Users can use `?CMGFM` to see the details about this function
```{r  eval = TRUE}
system.time({
  tic <- proc.time()
  rlist <- CMGFM(XList, Z, types=types, q=q, numvarmat=numvarmat)
  toc <- proc.time()
  time_smgfm <- toc[3] - tic[3]
})
```


Check the increased property of the evidence lower bound function.
```{r  eval = TRUE}
library(ggplot2)
dat_iter <- data.frame(iter=1:length(rlist$ELBO_seq), ELBO=rlist$ELBO_seq)
ggplot(data=dat_iter, aes(x=iter, y=ELBO)) + geom_line() + geom_point() + theme_bw(base_size = 20)

```

We calculate the metrics to measure the estimation accuracy, where the trace statistic is used to measure the estimation accuracy of loading matrix and prediction accuracy of factor matrix, which is evaluated by the function `measurefun()` in the R package `GFM`, and the root of mean square error is adopted to measure the estimation error of bbeta.
```{r  eval = TRUE}
library(GFM)
hUplist <- lapply(seq_along(rlist$Bf), function(m) cbind(rlist$muf[[m]], rlist$Bf[[m]]))
metricList <- list()
metricList$CMGFM <- list()
meanTr <- function(hBlist, Blist,  type='trace_statistic'){
  ###It is noted that the trace statistics is not symmetric, the true value must be in last
  trvec <- sapply(1:length(Blist), function(j) measurefun(hBlist[[j]], Blist[[j]], type = type))
  return(mean(trvec))
}
normvec <- function(x) sqrt(sum(x^2/ length(x)))
metricList$CMGFM$Tr_F <- measurefun(rlist$M, datlist$F0)
metricList$CMGFM$Tr_Upsilon <- meanTr(hUplist, Uplist)
metricList$CMGFM$Tr_U <- measurefun(Reduce(cbind,rlist$Xif), Reduce(cbind, datlist$U0))
metricList$CMGFM$beta_norm <- normvec(as.vector(Reduce(cbind,rlist$betaf)- Reduce(cbind,datlist$beta0List)))
metricList$CMGFM$Time <- rlist$time_use
```


## Compare with other methods

We compare CMGFM with various prominent methods in the literature. They are (1) High-dimensional LFM (Bai and Ng 2002) implemented in the R package GFM; (2) PoissonPCA (Kenney et al. 2021) implemented in the R package PoissonPCA; (3) Zero-inflated Poisson factor model (ZIPFA, Xu et al. 2021) implemented in the R package ZIPFA; (4) Generalized factor model (Liu et al. 2023) implemented in the R package GFM; (5) PLNPCA (Chiquet et al. 2018) implemented in the R package PLNmodels;  (6) Generalized Linear Latent Variable Models (GLLVM, Hui et al. 2017) implemented in the R package gllvm. (7) Poisson regression model for each $x_{ij}, (j = 1,··· ,p)$, implemented in stats R package; (8) Multi-response reduced-rank Poisson regression model (MMMR, Luo et al. 2018) implemented in rrpack R package.

(1). First, we implemented the generalized factor model (LFM) and record the metrics that measure the estimation accuracy and computational cost.

```{r  eval = TRUE}
metricList$GFM <- list()
tic <- proc.time()
res_gfm <- gfm(XList, types=types, q=q)
toc <- proc.time()
time_gfm <- toc[3] - tic[3]
mat2list <- function(B, pvec, by_row=TRUE){
  Blist <- list()
  pcum = c(0, cumsum(pvec))
  for(i in 1:length(pvec)){
    if(by_row){
      Blist[[i]] <- B[(pcum[i]+1):pcum[i+1],]
    }else{
      Blist[[i]] <- B[, (pcum[i]+1):pcum[i+1]]
    }
  }
  return(Blist)
}
metricList$GFM$Tr_F <- measurefun(res_gfm$hH, datlist$F0)
metricList$GFM$Tr_Upsilon <- meanTr(mat2list(cbind(res_gfm$hmu,res_gfm$hB), pvec), Uplist)
metricList$GFM$Tr_U <- NA
metricList$GFM$beta_norm <- NA
metricList$GFM$Time <- time_gfm
```

(2) Eightly, we implemented the first version of multi-response reduced-rank Poisson regression model (MMMR, Luo et al. 2018) implemented in rrpack R package (MRRR-Z), that did not consider the latent factor structure but only the covariates.

```{r  eval = TRUE}
mrrr_run <- function(Y, Z, numvarmat, rank0,family=list(poisson()),
                     familygroup,  epsilon = 1e-4, sv.tol = 1e-2,
                     lambdaSVD=0.1, maxIter = 2000, trace=TRUE, trunc=500){
  # epsilon = 1e-4; sv.tol = 1e-2; maxIter = 30; trace=TRUE,lambdaSVD=0.1
  Diag <- function(vec){
  q <- length(vec)
  if(q > 1){
    y <- diag(vec)
  }else{
    y <- matrix(vec, 1,1)
  }
  return(y)
}
  require(rrpack)
  q <- rank0
  n <- nrow(Y); p <- ncol(Y)
  X <- cbind(cbind(1, Z),  diag(n))
  d <- ncol(Z)
  ## Trunction
  Y[Y>trunc] <- trunc
  Y[Y< -trunc] <- -trunc
  
  tic <- proc.time()
  pvec <- as.vector(t(numvarmat))
  pvec <- pvec[pvec>0]
  pcums <- cumsum(pvec)
  idxlist <- list()
  idxlist[[1]] <- 1:pcums[1]
  if(length(pvec)>1){
    for(i in 2:length(pvec)){
      idxlist[[i]] <- (pcums[i-1]+1):pcums[i] 
    }
  }
  
  svdX0d1 <- svd(X)$d[1]
  init1 = list(kappaC0 = svdX0d1 * 5)
  offset = NULL
  control = list(epsilon = epsilon, sv.tol = sv.tol, maxit = maxIter,
                 trace = trace, gammaC0 = 1.1, plot.cv = TRUE,
                 conv.obj = TRUE)
  res_mrrr <- mrrr(Y=Y, X=X[,-1], family = family, familygroup = familygroup,
                   penstr = list(penaltySVD = "rankCon", lambdaSVD = lambdaSVD),
                   control = control, init = init1, maxrank = rank0+d) #
  
  hmu <- res_mrrr$coef[1,]
  hbeta <- t(res_mrrr$coef[2:(d+1),])
  hTheta <- res_mrrr$coef[-c(1:(d+1)),]
  
  hbeta_rf <- NULL
  for(i in seq_along(pvec)){
    hbeta_rf <- cbind(hbeta_rf, colMeans(hbeta[idxlist[[i]],]))
  }
  
  # Matrix::rankMatrix(hTheta)
  svd_Theta <- svd(hTheta, nu=q, nv=q)
  hH <- svd_Theta$u
  hB <- svd_Theta$v %*% Diag(svd_Theta$d[1:q])
  toc <- proc.time()
  time_mrrr <- toc[3] - tic[3]
  
  return(list(hH=hH, hB=hB, hmu= hmu, beta=hbeta_rf, time_use=time_mrrr))
}

family_use <- list(gaussian(), poisson(), binomial())
familygroup <- sapply(seq_along(datlist$XList), function(j) rep(j, ncol(datlist$XList[[j]])))
res_mrrr <- mrrr_run(Reduce(cbind, datlist$XList),  Z = Z, numvarmat, rank0=q, family=family_use,
                       familygroup = unlist(familygroup), maxIter=100) #
  



```
```{r  eval = TRUE}
metricList$MRRR <- list()
metricList$MRRR$Tr_F <- measurefun(res_mrrr$hH, datlist$F0)
metricList$MRRR$Tr_Upsilon <-meanTr(mat2list(cbind(res_mrrr$hmu,res_mrrr$hB), pvec), Uplist)
metricList$MRRR$Tr_U <-NA
metricList$MRRR$beta_norm <- normvec(as.vector(res_mrrr$beta- Reduce(cbind,datlist$beta0List)))
metricList$MRRR$Time <- res_mrrr$time_use
```



## Visualize the comparison of performance

Next, we summarized the metrics for COAP and other compared methods in a dataframe object.
```{r  eval = TRUE}
list2vec <- function(xlist){
  nn <- length(xlist)
  me <- rep(NA, nn)
  idx_noNA <- which(sapply(xlist, function(x) !is.null(x)))
  for(r in idx_noNA) me[r] <- xlist[[r]]
  return(me)
}

dat_metric <- data.frame(Tr_F = sapply(metricList, function(x) x$Tr_F), 
                         Tr_Upsilon = sapply(metricList, function(x) x$Tr_Upsilon),
                         Tr_U = sapply(metricList, function(x) x$Tr_U),
                         beta_norm =sapply(metricList, function(x) x$beta_norm),
                         Time = sapply(metricList, function(x) x$Time),
                         Method = names(metricList))
dat_metric
```

Plot the results for COAP and other methods, which suggests that CMGFM achieves better estimation accuracy for the quantities of interest.
```{r  eval = TRUE, fig.width=9, fig.height=6}
library(cowplot)
p1 <- ggplot(data=subset(dat_metric, !is.na(Tr_F)), aes(x= Method, y=Tr_F, fill=Method)) + geom_bar(stat="identity") + xlab(NULL) + scale_x_discrete(breaks=NULL) + theme_bw(base_size = 16)
p2 <- ggplot(data=subset(dat_metric, !is.na(Tr_Upsilon)), aes(x= Method, y=Tr_Upsilon, fill=Method)) + geom_bar(stat="identity") + xlab(NULL) + scale_x_discrete(breaks=NULL)+ theme_bw(base_size = 16)
p3 <- ggplot(data=subset(dat_metric, !is.na(beta_norm)), aes(x= Method, y=beta_norm, fill=Method)) + geom_bar(stat="identity") + xlab(NULL) + scale_x_discrete(breaks=NULL)+ theme_bw(base_size = 16)
p4 <- ggplot(data=subset(dat_metric, !is.na(Time)), aes(x= Method, y=Time, fill=Method)) + geom_bar(stat="identity") + xlab(NULL) + scale_x_discrete(breaks=NULL)+ theme_bw(base_size = 16)
plot_grid(p1,p2,p3, p4, nrow=2, ncol=2)
```


## Select the parameters

We applied the maximum singular value ratio based method to select the number of factors. The results showed that  the maximum SVR method has the potential to identify the true values.
```{r  eval = TRUE}

hq <- MSVR(XList, Z, types=types, numvarmat, q_max=20)

print(c(q_true=q, q_est=hq))

```


<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>